% future/HTMtableFull.tex
% SPDX-License-Identifier: CC-BY-SA-3.0

\begin{sidewaystable*}[htbp]
\input{future/HTMtableColor}
\centering
\caption{Comparison of Locking (Plain and Augmented) and HTM
  (\colorbox{plus}{Advantage}, \colorbox{minus}{Disadvantage},
  \colorbox{down}{Strong Disadvantage})}
\label{tab:future:Comparison of Locking (Plain and Augmented) and HTM}
\footnotesize
\setstretch{0.95}
%\renewcommand*{\arraystretch}{1.4}
\setlength{\tabcolsep}{3pt}
\resizebox{8in}{!}{
\begin{tabularx}{8.5in}{p{.85in}cXcXcX}
\toprule
  &
    & \multicolumn{1}{c}{Locking} &
      & \multicolumn{1}{c}{Locking with Userspace RCU or Hazard Pointers} &
        & \multicolumn{1}{c}{Hardware Transactional Memory} \\
\midrule
  Basic Idea &
    & Allow only one thread at a time to access a given set of objects. &
      & Allow only one thread at a time to access a given set of objects. &
        & Cause a given operation over a set of objects to execute atomically. \\
\midrule
  Scope &
    & \Pl Handles all operations. &
      & \Pl Handles all operations. &
        & \Pl Handles revocable operations. \\
\addlinespace[4pt]
  &
    & &
      & &
        & \Mn Irrevocable operations force fallback (typically to locking). \\
\midrule
  Composability &
    & \Dw Limited by deadlock. &
      & \Pl Readers limited only by grace-period-wait operations. &
        & \Dw Limited by irrevocable operations, transaction size, and deadlock.
          (Assuming lock-based fallback code.) \\
\addlinespace[4pt]
  &
    & &
      & \Mn Updaters limited by deadlock. Readers reduce deadlock. &
        & \\
\midrule
  Scalability \& Performance &
    & \Mn Data must be partitionable to avoid lock contention. &
      & \Mn Data must be partitionable to avoid lock contention among updaters. &
        & \Mn Data must be partitionable to avoid conflicts. \\
\addlinespace[4pt]
  &
    & &
      & \Pl Partitioning not needed for readers. &
        & \\
\cmidrule{3-7}
  &
    & \Dw Partitioning must typically be fixed at design time. &
      & \Dw Partitioning for updaters must typically be fixed at design time. &
        & \Pl Dynamic adjustment of partitioning carried out automatically
          down to cacheline boundaries. \\
\addlinespace[4pt]
  &
    & &
      & \Pl Partitioning not needed for readers. &
        & \Mn Partitioning required for fallbacks (less important for rare
          fallbacks). \\
\cmidrule{3-7}
  &
    & \Dw Locking primitives typically result in expensive cache misses and
      memory-barrier instructions.&
      & \Dw Updater locking primitives typically result in expensive cache
        misses and memory-barrier instructions. &
        & \Mn Transactions begin/end instructions typically do not result in
          cache misses, but do have memory-ordering consequences. \\
\cmidrule{3-7}
  &
    & \Pl Contention effects are focused on acquisition and release, so that
      the critical section runs at full speed. &
      & \Pl Update-side contention effects are focused on acquisition and
        release, so that the critical section runs at full speed. &
        & \Mn Contention aborts conflicting transactions, even if they have been
          running for a long time. \\
\addlinespace[4pt]
  &
    & &
      & \Pl Readers do not contend with updaters or with each other. &
        & \\
\cmidrule{3-7}
  &
    & &
      & \Pl Read-side primitives are typically wait-free with low overhead.
        (Lock-free for hazard pointers.) &
        & \Mn Read-only transactions subject to conflicts and rollbacks. No
          forward-progress guarantees other than those supplied by fallback
          code. \\
\cmidrule{3-7}
  &
    & \Pl Privatization operations are simple, intuitive, performant,
      and scalable. &
      & \Pl Privatization operations are simple, intuitive, performant,
        and scalable when data is visible only to updaters. &
        & \Mn Privatized data contributes to transaction size. \\
\addlinespace[4pt]
  &
    & &
      & \Mn Privatization operations are expensive (though still intuitive
        and scalable) for reader-visible data. &
        & \\
\midrule
  Hardware Support &
    & \Pl Commodity hardware suffices. &
      & \Pl Commodity hardware suffices. &
        & \Mn New hardware required (and is starting to become available). \\
\cmidrule{3-7}
  &
    & \Pl Performance is insensitive to cache-geometry details. &
      & \Pl Performance is insensitive to cache-geometry details. &
        & \Mn Performance depends critically on cache geometry. \\
\midrule
  Software Support &
    & \Pl APIs exist, large body of code and experience, debuggers operate
      naturally. &
      & \Pl APIs exist, large body of code and experience, debuggers operate
        naturally. &
        & \Mn APIs emerging, little experience outside of DBMS, breakpoints
              mid-transaction can be problematic. \\
\midrule
  Interaction With Other Mechanisms &
    & \Pl Long experience of successful interaction. &
      & \Pl Long experience of successful interaction. &
        & \Dw Just beginning investigation of interaction. \\
\midrule
  Practical Apps &
    & \Pl Yes. &
      & \Pl Yes. &
        & \Pl Yes. \\
\midrule
  Wide Applicability &
    & \Pl Yes. &
      & \Pl Yes. &
        & \Mn Jury still out, but likely to win significant use. \\
\bottomrule
\end{tabularx}
}
\end{sidewaystable*}
