% together/hash.tex
% SPDX-License-Identifier: CC-BY-SA-3.0

\section{Hashing Hassles}
\label{sec:together:Hashing Hassles}

This section looks at some issues that can arise when dealing with
hash tables.
Please note that these issues also apply to many other search structures.

\subsection{Correlated Data Elements}
\label{sec:together:Correlated Data Elements}

This situation is analogous to that in
Section~\ref{sec:together:Correlated Fields}:
We have a hash table where we need correlated views of two or more of
the elements.
These elements are updated together, and we do not want to see an old
version of the first element along with new versions of the other
elements.
For example, Schr\"odinger decided to add his extended family to his
in-memory database along with all his animals.
Although Schr\"odinger understands that marriages and divorces do not
happen instantaneously, he is also a traditionalist.
As such, he absolutely does not want his database ever to show that the
bride is now married, but the groom is not, and vice versa.
In other words, Schr\"odinger wants to be able to carry out a
wedlock-consistent traversal of his database.

One approach is to use sequence locks
(see Section~\ref{sec:defer:Sequence Locks}),
so that wedlock-related updates are carried out under the
protection of \co{write_seqlock()}, while reads requiring
wedlock consistency are carried out within
a \co{read_seqbegin()} / \co{read_seqretry()} loop.
Note that sequence locks are not a replacement for RCU protection:
Sequence locks protect against concurrent modifications, but RCU
is still needed to protect against concurrent deletions.

This approach works quite well when the number of correlated elements is
small, the time to read these elements is short, and the update rate is
low.
Otherwise, updates might happen so quickly that readers might never complete.
Although Schr\"odinger does not expect that even his least-sane relatives
will marry and divorce quickly enough for this to be a problem,
he does realize that this problem could well arise in other situations.
One way to avoid this reader-starvation problem is to have the readers
use the update-side primitives if there have been too many retries,
but this can degrade both performance and scalability.

In addition, if the update-side primitives are used too frequently,
poor performance and scalability will result due to lock contention.
One way to avoid this is to maintain a per-element sequence lock,
and to hold both spouses' locks when updating their marital status.
Readers can do their retry looping on either of the spouses' locks
to gain a stable view of any change in marital status involving both
members of the pair.
This avoids contention due to high marriage and divorce rates, but
complicates gaining a stable view of all marital statuses during a
single scan of the database.

If the element groupings are well-defined and persistent, which marital
status is hoped to be,
then one approach is to add pointers to the data elements to link
together the members of a given group.
Readers can then traverse these pointers to access all the data elements
in the same group as the first one located.

Other approaches using version numbering are left as exercises for the
interested reader.

\subsection{Update-Friendly Hash-Table Traversal}
\label{sec:together:Update-Friendly Hash-Table Traversal}

Suppose that a statistical scan of all elements in a hash table is
required.
For example, Schr\"odinger might wish to compute the average
length-to-weight ratio over all of his animals.\footnote{
	Why would such a quantity be useful?
	Beats me!
	But group statistics in general are often useful.}
Suppose further that Schr\"odinger is willing to ignore slight
errors due to animals being added to and removed from the hash
table while this statistical scan is being carried out.
What should Schr\"odinger do to control concurrency?

One approach is to enclose the statistical scan in an RCU read-side
critical section.
This permits updates to proceed concurrently without unduly impeding
the scan.
In particular, the scan does not block the updates and vice versa,
which allows scan of hash tables containing very large numbers of
elements to be supported gracefully, even in the face of very high
update rates.

\QuickQuiz{}
	But how does this scan work while a resizable hash table
	is being resized?
	In that case, neither the old nor the new hash table is
	guaranteed to contain all the elements in the hash table!
\QuickQuizAnswer{
	True, resizable hash tables as described in
	Section~\ref{sec:datastruct:Non-Partitionable Data Structures}
	cannot be fully scanned while being resized.
	One simple way around this is to acquire the
	\co{hashtab} structure's \co{->ht_lock} while scanning,
	but this prevents more than one scan from proceeding
	concurrently.

	Another approach is for updates to mutate the old hash
	table as well as the new one while resizing is in
	progress.
	This would allow scans to find all elements in the old
	hash table.
	Implementing this is left as an exercise for the reader.
} \QuickQuizEnd
