% appendix/questions/questions.tex
% mainfile: ../../perfbook.tex
% SPDX-License-Identifier: CC-BY-SA-3.0

\QuickQuizChapter{cha:app:Important Questions}{Important Questions}{qqzquestions}
%
\Epigraph{Ask me no questions, and I'll tell you no fibs.}
	 {``She Stoops to Conquer'', Oliver Goldsmith}

The following sections discuss some important questions relating to
SMP programming.
Each section also shows how to avoid worrying about
the corresponding question, which can be extremely important if
your goal is to simply get your SMP code working as quickly and
painlessly as possible---which is an excellent goal, by the way!

Although the answers to these questions are often less
intuitive than they would be in a single-threaded setting,
with a bit of work, they are not that difficult to understand.
If you managed to master recursion, there is nothing here that should
pose an overwhelming challenge.

With that, here are the questions:

\begin{enumerate}
\item	Why aren't parallel programs always faster?
	(\cref{sec:app:questions:Why Aren't Parallel Programs Always Faster?})
\item	Why not remove locking?
	(\cref{sec:app:questions:Why Not Remove Locking?})
\item	What time is it?
	(\cref{sec:app:questions:What Time Is It?})
\item	What does ``after'' mean?
	(\cref{sec:app:questions:What Does ``After'' Mean?})
\item	How much ordering is needed?
	(\cref{sec:app:questions:How Much Ordering Is Needed?})
\item	What is the difference between ``concurrent'' and ``parallel''?
	(\cref{sec:app:questions:What is the Difference Between ``Concurrent'' and ``Parallel''?})
\item	Why is software buggy?
	(\cref{sec:app:questions:Why Is Software Buggy?})
\end{enumerate}

Read on to learn some answers.
Improve upon these answers if you can!

\input{appendix/questions/parallelfaster}
\input{appendix/questions/removelocking}
\input{appendix/questions/time}
\input{appendix/questions/after}
\input{appendix/questions/ordering}
\input{appendix/questions/concurrentparallel}
\input{appendix/questions/buggy}

% @@@ How much documentation is enough?

\QuickQuizAnswersChp{qqzquestions}
