% legal.tex
% mainfile: perfbook.tex
% SPDX-License-Identifier: CC-BY-SA-3.0

\section*{Legal Statement}

This work represents the views of the editor and the authors and does not
necessarily represent the view of their respective employers. \\
\\
Trademarks:
\begin{itemize}
\item	IBM, z~Systems, and PowerPC are trademarks or registered trademarks
	of International Business Machines Corporation in the United
	States, other countries, or both.
\item	Linux is a registered trademark of Linus Torvalds.
\item	Intel, Itanium, Intel Core, and Intel Xeon are trademarks
	of Intel Corporation or its subsidiaries in the United States,
	other countries, or both.
\item	Arm is a registered trademark of Arm Limited (or its subsidiaries)
	in the US and/or elsewhere.
\item	SPARC is a registered trademark of SPARC International, Inc.
	Products bearing SPARC trademarks are based on an architecture
	developed by Sun Microsystems, Inc.
\item	Other company, product, and service names may be trademarks or
	service marks of such companies.
\end{itemize}

The non-source-code text and images in this document are provided under
the terms of the Creative Commons Attribution-Share Alike 3.0 United
States license.\footnote{
	\url{https://creativecommons.org/licenses/by-sa/3.0/us/}}
In brief, you may use the contents of this document for any purpose,
personal, commercial, or otherwise, so long as attribution to the
authors is maintained.  Likewise, the document may be modified,
and derivative works and translations made available, so long as
such modifications and derivations are offered to the public on equal
terms as the non-source-code text and images in the original document.

Source code is covered by various versions of the GPL\@.\footnote{
	\url{https://www.gnu.org/licenses/gpl-2.0.html}}
Some of this code is GPLv2-only, as it derives from the Linux kernel,
while other code is GPLv2-or-later.
See the comment headers of the individual source files within the
CodeSamples directory in the git archive\footnote{
	\url{git://git.kernel.org/pub/scm/linux/kernel/git/paulmck/perfbook.git}}
for the exact licenses.
If you are unsure of the license for a given code fragment,
you should assume GPLv2-only.

Combined work {\textcopyright}~2005--\commityear\ by Paul E. McKenney.
Each individual contribution is copyright by its contributor at the time
of contribution, as recorded in the git archive.
