% defer/rcu.tex
% mainfile: ../perfbook.tex
% SPDX-License-Identifier: CC-BY-SA-3.0

\section{Read-Copy Update (RCU)}
\label{sec:defer:Read-Copy Update (RCU)}
%
\epigraph{``Free'' is a \emph{very} good price!}{Tom Peterson}

All of the mechanisms discussed in the preceding sections
used one of a number of approaches to defer specific actions
until they may be carried out safely.
The reference counters discussed in
\cref{sec:defer:Reference Counting}
use explicit counters to defer actions that could disturb readers,
which results in read-side contention and thus poor scalability.
The hazard pointers covered by
\cref{sec:defer:Hazard Pointers}
uses implicit counters in the guise of per-thread lists of pointer.
This avoids read-side contention, but requires readers to do stores and
conditional branches, as well as either \IXhpl{full}{memory barrier}
in read-side primitives or real-time-unfriendly \IXacrlpl{ipi} in
update-side primitives.\footnote{
	In some important special cases, this extra work can be avoided
	by using link counting as exemplified by the UnboundedQueue
	and ConcurrentHashMap data structures implemented in Folly
	open-source library
	(\url{https://github.com/facebook/folly}).}
The sequence lock presented in
\cref{sec:defer:Sequence Locks}
also avoids read-side contention, but does not protect pointer
traversals and, like hazard pointers, requires either full memory barriers
in read-side primitives, or inter-processor interrupts in update-side
primitives.
These schemes' shortcomings raise the question of
whether it is possible to do better.

This section introduces \IXBacrfst{rcu}, which provides
an API that allows readers to be associated with regions in the source code,
rather than with expensive updates to frequently updated shared data.
The remainder of this
section examines RCU from a number of different perspectives.
\Cref{sec:defer:Introduction to RCU} provides the classic
introduction to RCU,
\cref{sec:defer:RCU Fundamentals} covers fundamental RCU
concepts,
\cref{sec:defer:RCU Linux-Kernel API} presents the Linux-kernel
API,
\cref{sec:defer:RCU Usage} introduces some common RCU use cases,
and finally
\cref{sec:defer:RCU Related Work} covers recent work related
to RCU.

Although RCU has gained a reputation for being subtle and difficult,
when used properly, it is quite straightforward.
In fact, no less an authority than Butler Lampson classifies it as easy
concurrency~\cite[Chapter 3]{ButlerLampson2022ProgrammingConcurrentSystems}.
And Lampson is quite right:
The core concept behind RCU is nothing more than that updaters are
guaranteed to wait for pre-existing (and \emph{only} pre-existing)
readers to complete.
For their part, Balmau, Guerraoui, Herlihy, and Zablotchi state
``QSBR [a particular class of RCU implementations] is fast and can be
applied to virtually any data structure''.

On the other hand, making good use of RCU often requires that you think
differently about concurrency in general and about your use case in
particular.
Much of the remainder of this section therofore provides a guide to
mapping use cases onto RCU.

\QuickQuiz{
	RCU cannot possibly be this simple!!!
	And what on earth could you possibly do with a synchronization
	primitive that is nothing more than a simple delay???
}\QuickQuizAnswer{
	I am sorry, but the core of RCU really is exactly that simple!

	And you can do many amazing things with RCU's simple delay.
	Please feel free to continue reading this section in order to
	learn how.
}\QuickQuizEnd

\input{defer/rcuintro}
\input{defer/rcufundamental}
\input{defer/rcuapi}
\input{defer/rcuusage}
\input{defer/rcurelated}
