% defer/rcurelated.tex

\subsection{RCU Related Work}
\label{sec:defer:RCU Related Work}
\OriginallyPublished{Section}{sec:defer:RCU Related Work}{RCU Related Work}{Linux Weekly News}{PaulEMcKenney2014ReadMostly}
\OriginallyPublished{Section}{sec:defer:RCU Related Work}{RCU Related Work}{Linux Weekly News}{PaulEMcKenney2015ReadMostly}

The known first mention of anything resembling RCU took the form of a bug
report from
Donald Knuth~\cite[page 413 of Fundamental Algorithms]{Knuth73}
against J. Weizenbaum's SLIP list-processing facility for
FORTRAN~\cite{Weizenbaum:1963:SLP:367593.367617}.
Knuth was justified in reporting the bug, as SLIP had no notion of
any sort of grace-period guarantee.

The first known non-bug-report mention of anything resembing RCU appeared
in Kung's and Lehman's landmark paper~\cite{Kung80}.
There was some additional use of this technique in
academia~\cite{Manber82,Manber84,BarbaraLiskov1988ArgusCACM,Pugh90,Andrews91textbook,Pu95a,Cowan96a,Rastogi:1997:LPV:645923.671017,Gamsa99},
but much of the work in this area was carried out by
practitioners~\cite{RichardRashid87a,Hennessy89,Jacobson93,AjuJohn95,Slingwine95,Slingwine97,Slingwine98,McKenney98}.
By the year 2000, the initiative had passed to open-source projects,
most notably the Linux kernel
community~\cite{RustyRussell2000a,RustyRussell2000b,McKenney01b,McKenney01a,McKenney02a,Arcangeli03}.\footnote{
	A list of citations with well over 200 entries may be found in
	\co{bib/RCU.bib} in the {\LaTeX} source for this book.}

However, in the mid 2010s, there was a welcome upsurge in RCU research
and development across a number of communities and institutions.
Section~\ref{sec:defer:RCU Uses} describes uses of RCU,
Section~\ref{sec:defer:RCU Implementations} describes RCU implementations
(as well as work that both creates and uses an implementation),
and finally,
Section~\ref{sec:defer:RCU Validation} describes verification and validation
of RCU and its uses.

\subsubsection{RCU Uses}
\label{sec:defer:RCU Uses}

Phil Howard and Jon Walpole of Portland State University (PSU) have
applied RCU to red-black
trees~\cite{PhilHowardPhD,PhilHoward2011RCUTMRBTree} combined with updates
synchronized using software transactional memory.
Josh Triplett and Jon Walpole (again of PSU) applied RCU to resizable
hash tables~\cite{JoshTriplettPhD,Triplett:2011:RPHash,JonCorbet2014RCUhash1,JonCorbet2014RCUhash2}.
Other RCU-protected resizable hash tables have been created by
Herbert Xu~\cite{HerbertXu2010RCUResizeHash} and by
Mathieu Desnoyers~\cite{PaulMcKenney2013LWNURCUhash}.

Austin Clements, Frans Kaashoek, and Nickolai Zeldovich
of MIT created an RCU-optimized balanced binary tree
(Bonsai)~\cite{AustinClements2012RCULinux:mmapsem}, and applied this
tree to the Linux kernel's VM subsystem in order to reduce read-side
contention on the Linux kernel's \co{mmap_sem}.
This work resulted in order-of-magnitude speedups and scalability up to
at least 80 CPUs for a microbenchmark featuring large numbers of minor
page faults.
This is similar to a patch developed earlier by
Peter Zijlstra~\cite{PeterZijlstra2014SpeculativePageFault}, and both
were limited by the fact that, at the time, filesystem data structures
were not safe for RCU readers.
Clements et al. avoided this limitation by optimizing the page-fault
path for anonymous pages only.
More recently, filesystem data structures have been made safe for RCU
readers~\cite{JonathanCorbet2010dcacheRCU,JonathanCorbet2011dcacheRCUbug},
so perhaps this work can be implemented for all page types, not just
anonymous pages---Peter Zijlstra has, in fact, recently prototyped
exactly this.

Yandong Mao and Robert Morris of MIT and Eddie Kohler of
Harvard University created another RCU-protected tree named
Masstree~\cite{Mao:2012:CCF:2168836.2168855} that combines ideas from B+
trees and tries.
Although this tree is about 2.5x slower than an RCU-protected hash table,
it supports operations on key ranges, unlike hash tables.
In addition, Masstree supports efficient storage of objects with long
shared key prefixes and, furthermore, provides persistence via logging
to mass storage.

The paper notes that Masstree's performance rivals that of memcached, even
given that Masstree is persistently storing updates and memcached is not.
The paper also compares Masstree's performance to the persistent
datastores MongoDB, VoltDB, and Redis, reporting significant performance
advantages for Masstree, in some cases exceeding two orders of magnitude.
Another paper~\cite{Tu:2013:STM:2517349.2522713}, by Stephen Tu,
Wenting Zheng, Barbara Liskov, and Samuel Madden of MIT and Kohler,
applies Masstree to an in-memory database named Silo, achieving 700K
transactions per second (42M transactions per minute) on a well-known
transaction-processing benchmark.
Interestingly enough, Silo guarantees linearizability without incurring
the overhead of grace periods while holding locks.

\subsubsection{RCU Implementations}
\label{sec:defer:RCU Implementations}

Mathieu Desnoyers created a user-space RCU for use in
tracing~\cite{MathieuDesnoyersPhD,MathieuDesnoyers2012URCU},
which has seen use in a number of projects~\cite{MikeDay2013RCUqemu}.

Maya Arbel and Hagit Attiya of Technion took a more rigorous
approach~\cite{MayaArbel2014RCUtree} to an RCU-protected search tree that,
like Masstree, allows concurrent updates.
This paper includes a proof of correctness, including proof that all
operations on this tree are linearizable.
Unfortunately, this implementation achieves linearizability by incurring
the full latency of grace-period waits while holding locks, which degrades
scalability of update-only workloads.
One way around this problem is to abandon
linearizability~\cite{AndreasHaas2012FIFOisnt,PaulEMcKennneyAtomicTreeN4037}),
however, Arbel and Attiya instead created an RCU variant that reduces
low-end grace-period latency.
Of course, nothing comes for free, and this RCU variant appears to hit
a scalability limit at about 32 CPUs.
Although there is much to be said for dropping linearizability, thus
gaining both performance and scalability, it is very good to see academics
experimenting with alternative RCU implementations.
Researchers at Charles University in Prague have also been
working on RCU implementations, including dissertations by
Andrej Podzimek~\cite{AndrejPodzimek2010masters} and Adam
Hraska~\cite{AdamHraska2013RCUHelenOS}.

RCU-like mechanisms are also finding their way into Java.
Sivaramakrishnan et al.~\cite{Sivaramakrishnan:2012:ERB:2258996.2259005}
use an RCU-like mechanism to eliminate the read barriers that are
otherwise required when interacting with Java's garbage collector,
resulting in significant performance improvements.

Ran Liu, Heng Zhang, and Haibo Chen of Shanghai Jiao Tong University
created a specialized variant of RCU that they used for an optimized
``passive reader-writer lock''~\cite{RanLiu2014PassiveRWLock}, similar to
those created by Gautham Shenoy~\cite{GauthamShenoy2006RCUrwlock} and
Srivatsa Bhat~\cite{SrivatsaSBhat2014RCUrwlock}.
The Liu et al.~paper is interesting from a number of
perspectives~\cite{PaulEMcKenney2014ReadMostly}.

Geoff Romer and Andrew Hunter proposed a cell-based API for RCU
protection of singleton data structures for inclusion in the
C++ standard~\cite{GeoffRomer2017C++DeferredReclamation}.

\subsubsection{RCU Validation}
\label{sec:defer:RCU Validation}

In early 2017, it is commonly recognized that almost any bug is a potential
security exploit, so validation and verification are first-class concerns.

Researchers at Stony Brook University have produced an RCU-aware data-race
detector~\cite{AbhinavDuggal2010Masters,JustinSeyster2012PhD,Seyster:2011:RFA:2075416.2075425}.
Alexey Gotsman of IMDEA, Noam Rinetzky of Tel Aviv University,
and Hongseok Yang of the University of Oxford have published a
paper~\cite{AlexeyGotsman2012VerifyGraceExtended} expressing the formal
semantics of RCU in terms of separation logic, and have continued with
other aspects of concurrency.

With some luck, all of this validation work will eventually result in
more and better tools for validating concurrent code.
